---
title: "Bayesian Clinical Trials"
subtitle: Informative vs. non-informative beta priors  
output:
  ioslides_presentation:
    css: assets/css/ioslides.css
    logo: assets/img/logos3.png
    mathjax: local
    self_contained: no
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
---

## Construction prior Beta
 
## Updatindg with Data
```{r}
library(LearnBayes)
```

Suppose we are interested in the response $p$ of a drug. 

- The  function __bayes.select__ allow for specifying a beta prior based on
knowledge of two prior quantiles. 

- Suppose the prior median for the response rate is 0.2 and the 75th percentile is 0.3.

```{r}
beta.prior = beta.select(list(p=0.5, x=0.2), list(p=0.75, x=.3))
print(beta.prior)
```

A beta(2.04, 7.19) prior matches this prior information

## Updating with data
- Next, suppose to observe for 3 successive patients no adverse events
      + 3 successes and 0 failures
      
The posterior distribution is...

## Triplot
The triplot function shows the prior, likelihood, and posterior on the same
display

```{r, fig.align='center'}
beta.prior = beta.select(list(p=0.5, x=0.2), list(p=0.75, x=.3))
triplot(beta.prior, c(3,0))
```

## Inference by sampling from the posterior
Inference about the response rate can be carried out by simulating a large number
of draws from the posterior and summarizing the simulated sample.

```{r}
beta.prior = beta.select(list(p=0.5, x=0.2), list(p=0.75, x=.3))
beta.post = beta.prior + c(3,0)
post.sample = rbeta(1000, beta.post[1], beta.post[2])
quantile(post.sample, c(0.05, 0.95))
```


## Predictive distribution

Suppose we want to predict the number of no adverse events (successes) in the next cohort of 3 patients.

```{r, eval=TRUE}

beta.prior = beta.select(list(p=0.5, x=0.2), list(p=0.75, x=.3))

n = 3
s = 0:n
pred.probs = pbetap(beta.prior, n, s)
discint(cbind(s, pred.probs), 0.95)
```


## Predictive distribution

```{r, fig.align='center', echo=FALSE}

beta.prior = beta.select(list(p=0.5, x=0.2), list(p=0.75, x=.3))

n = 3
s = 0:n
pred.probs = pbetap(beta.prior, n, s)
plot(s, pred.probs, type="h")
```



## Getting the slides

* The slides for this course were created with Rmarkdown: [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/). 
* They are available from [https://github.com/berkeley3/bayesianCT-course](https://github.com/berkeley3/bayesianCT-course).
* To re-compile the slides:

    + Download the directory containing the lecture from Github
    + In R open the .Rmd file and set the working directory to the lecture directory
    + Click the *KnitHTML* button on Rstudio or run the following commands: 
  
```{r RmarkdownChunk, eval=FALSE}
library(rmarkdown) 
render("index.Rmd")
```

