---
title: "Bayesian Clinical Trials"
subtitle: Case study  
output:
  ioslides_presentation:
    css: assets/css/ioslides.css
    logo: assets/img/logos3.png
    mathjax: local
    self_contained: no
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
---

## Example
Osteosarcoma is the most common malignant bone tumor in children and young adults and prognosis after its recurrence is poor. 

There is a need to identify novel effective agents and these agents are usually evaluated initially in patients with relapsed osteosarcoma.

Consider the design of a controlled phase II B trial to compare two different treatment, say A and B, with (dichotomized) response measured according to RECIST criteria as outcome. 

What sample size is needed to provide sufficient information to specify the true difference in response rates to within a total interval width of 20 percentage points? 

## Evidence

| Phase II Study | Treatment | Sample Size | Responses (%)  |95% CI        |
| -------------- | --------- | ----------- | -------------- |------------- |
| 1              | A         | 26          | 5 (19)         |              |
| 2              | B         | 37          | 14 (37)        | 22;53        |
| 3              | A         | 14          | 4 (28.5)       |              |

Assume a pooled estimate for Treatment A: $\frac{9}{40}=22.5$

Assume an estimate for Treatment B: $\frac{14}{37}=38.0$

## Frequentist Sample size (per group) by $l$ and $(1-\alpha)$  ##
If we consider not only the estimates but also uncertainty, for example the upper and lower limit of the 95% confidence interval, we can get different sample sizes:
```{r,  eval=F}
library(SampleSizeProportions)
```
```{r,echo=F}
suppressWarnings(library(SampleSizeProportions))
len<-c(0.20,0.21,0.22,0.24,0.25)
level<-0.95
p1<-0.23
p2<-c(0.22,0.25,0.38,0.53)
a<-matrix(0,length(len),length(p2))
colnames(a)<-paste("theta=",p2-p1)
rownames(a)<-paste("length=",len)
for (i in 1:length(len)) {for (j in 1:length(p2))
{a[i,j]<-propdiff.freq(len=len[i], p1,p2[j], level = level)[1]}
                        }
print(a)
```
Which is best?

## Priors ##
* For Treatment A we can assume that 0.225 is the mean of a Beta distribution: $Beta\left(c_{1}=9,d_{1}=31\right)$
* For Treatment B we can assume that 0.38 is the mean of a Beta distribution: $Beta\left(c_{2}=14,d_{2}=23\right)$

```{r,echo=F}
suppressWarnings(library(SampleSizeProportions))
x<-rbeta(10000,9,31)
curve(dbeta(x,9,31),xlab="Prob of response",ylab="Beta densities")
x<-rbeta(10000,14,23)
curve(dbeta(x,14,23),xlab="Prob of response",add=T,col="red",ylab="Beta densities")
legend(0.7,5, c("Treatment A","Treatmet B"), lty=c(1,1), col =c("black","red"),bty="n")
```

## SampleSizeProportions package
SampleSizeProportions implements a set of functions for calculating required sample sizes for 

* Average Length Criterion (ALC)
* Average Coverage Criterion (ACC)
* Worst Outcome Criterion (WOC) 

in the context of binomial observations.

```{r,eval=F}
library(SampleSizeProportions)
```
## Bayesian Sample size (per group) by $l$ and $(1-\alpha)$  ##

```{r,echo=F}
library(SampleSizeProportions)
len<-c(0.20,0.22,0.25)
level<-c(0.90,0.95,0.99)
a<-matrix(0,length(len),length(level))
b<-c<-a
colnames(a)<-paste("lev=",level)
rownames(a)<-paste("length=",len)
rownames(b)<-rownames(c)<-rownames(a)
for (i in 1:length(len)) {for (j in 1:length(level))
{a[i,j]<-propdiff.acc(len=len[i], 9, 31, 14, 23, level = level[j], equal = TRUE, m = 10000, mcs = 3)[1]
 b[i,j]<-propdiff.alc(len=len[i], 9, 31, 14, 23, level = level[j], equal = TRUE, m = 10000, mcs = 3)[1] 
 c[i,j]<-propdiff.woc(len=len[i], 9, 31, 14, 23, level = level[j], equal = TRUE)[1]}      }
print(rbind("ACC",a,"-","ALC",b,"-","WOC",c),quote=F)
```







